---
title: "Final report"
author: "Rob Arbon"
date: "09/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The Old Faithful data set contains the waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA. 

```{r load_data}
df <- faithful
```

The data description is given in the table below. 

| variable | description | 
| - | - |
| `eruptions` | the duration of the eruption, minutes |
| `waiting` | waiting time to next eruption, minutes | 


A  summary of the distributions of each variable are as follows:

```{r summary}
summary(df)
```

The marginal distributions of the variables shown below: 

```{r distribution}
hist(df$eruptions)
hist(df$waiting)
```

Both the duration of eruptions and waiting time show a bimodal distribution. This prompts the hypothesis: 

*There are 2 distinct types of eruption from Old Faithful: 1) short frequent eruptions, 2) long infrequent eruptions.*


# Methods

We will test this using a Gaussian mixture model to cluster the data into 1 and 2 clusters. 
The true number of clusters will correspond to the model with the smallest value of the integrated complete data likelihood (ICL).  

# Results

The R package `Rmixmod` was used to estimate two Gaussian mixture models. 

```{r}
library(Rmixmod)
# mixmodCluster fits a Gaussian mixture model
# nbCluster is the number of mixture components. 
mod1 <- mixmodCluster(faithful, nbCluster=1, criterion="ICL")
mod2 <- mixmodCluster(faithful, nbCluster=2, criterion="ICL")
```

A summary of both models is shown below. 
```{r}
summary(mod1)
summary(mod2)
```

The model with 1 component has an ICL value or `r mod1@bestResult@criterionValue`

The model with 2 components has an ICL value of `r mod2@bestResult@criterionValue`. 

The best performing model has `r if (mod1@bestResult@criterionValue < mod2@bestResult@criterionValue) print(1) else (2)` components. 

A plot of the optimal clustering is shown below. 

```{r}

plot(mod2)

```

# Conclusion

There do appear to be two different types of eruption.

